Refactoring the \texttt{nls()} function in R

### 2022-07-27 RefactoringNLS/RefactoringNLS.Rmd
Some cleanup of bib file
Edits to formula section

Specifying a model to a solver should, ideally, use the same syntax across
solver tools. Unfortunately, R allows multiple approaches within different
modelling tools, and within `nls()` itself.

@@

Within `nls()`, consider the following FOUR different specifications for the same problem, plus 
an intuitive choice, labelled fm2a, that does not work. In this failed
attempt, putting the `Asym` parameter in the model causes the `plinear` algorithm
to try to add another term to the model. We believe this is unfortunate, and would
like to see a consistent syntax. At the time of writing, we do
not foresee a resolution for this issue. In the example, we have NOT evaluated
the commands to save space.

@@

@@

<!-- Here we use a truncated version of the example to save display space. -->

<!-- ```{r nlsindx1} -->
<!-- ## The muscle dataset in MASS is from an experiment on muscle -->
<!-- ## contraction on 21 animals.  The observed variables are Strip -->
<!-- ## (identifier of muscle), Conc (Cacl concentration) and Length -->
<!-- ## (resulting length of muscle section). -->
<!-- if(! requireNamespace("MASS", quietly = TRUE)) stop("Need MASS pkg") -->
<!-- mm<- MASS::muscle[1:12,] # take only 1st few values of Strip -->
<!-- str(mm) -->
<!-- mm<-droplevels(mm) -->
<!-- str(mm) -->
<!-- nlev <- nlevels(mm) -->
<!-- withAutoprint({ -->
<!--   ## The non linear model considered is -->
<!--   ##       Length = alpha + beta*exp(-Conc/theta) + error -->
<!--   ## where theta is constant but alpha and beta may vary with Strip. -->
<!--   with(mm, table(Strip)) # 2, 3 or 4 obs per strip -->
<!--   nl <- nlevels(mm$Strip) -->
<!--   ## We first use the plinear algorithm to fit an overall model, -->
<!--   ## ignoring that alpha and beta might vary with Strip. -->
<!--   musc.1 <- nls(Length ~ cbind(1, exp(-Conc/th)), mm, -->
<!--                 start = list(th = 1), algorithm = "plinear") -->
<!--   summary(musc.1) -->

<!--   ## Then we use nls' indexing feature for parameters in non-linear -->
<!--   ## models to use the conventional algorithm to fit a model in which -->
<!--   ## alpha and beta vary with Strip.  The starting values are provided -->
<!--   ## by the previously fitted model. -->
<!--   ## Note that with indexed parameters, the starting values must be -->
<!--   ## given in a list (with names): -->
<!--   ## ?? but why use b here AND in the new formula?? -->
<!--   b <- coef(musc.1) -->

<!--   musc.2 <- nls(Length ~ a[Strip] + b[Strip]*exp(-Conc/th), data=mm, -->
<!--                 start = list(a = rep(b[2], nl), b = rep(b[3], nl), th = b[1])) -->
<!--   summary(musc.2) -->
<!-- }) -->

Running the example reveals that the answers for the parameters are NOT indexed. 
<!-- e.g. `coef(musc.2)` is a single level vector of parameters.  -->
We do not see `a[1], a[2], a[3]` but `a1, a2, a3`. 
<!-- This is because the model -->
<!-- must integrate all the parameters because `th` is a common parameter across the index `Strip`. -->
This is no doubt because programming for indexed parameters is extremely challenging.
However, we believe the current structure is could cause confusion and error, and 
propose an alternative approach below.

@@

- to provide a packaged version of `nls()` (call it `nlsalt`) coded entirely in R 
  that matches the features in base R or what is packaged in `nlspkg` as described 
  in @PkgFromRbase22.

@@

- to explain what we do, either in comments or separate maintainer documentation.
  Since we are complaining about the lack of explanatory material for the current
  code, we feel it is incumbent on us to provide such material for our own work, and if
  possible for the existing code.

@@

While not consistent with the `lm()` specification of linear models, 
nonlinear models need all parameters specified. 
That is, the model formula should be written down as it would be computed.
Thus a linear model (which is valid input for nonlinear estimatin) should
be specified `y ~ a * x + b` rather than `y ~ x`. 
Moreover, we want to avoid the partially linear parameters appearing 
in the result object as `.lin1`, `.lin2`, etc., where the actual position 
of parameters in the model is not explicit.

>>>>

The variable projection method (@Golub1973, @OlearyRust13) is generally much more 
effective then general approaches in finding good solutions to nonlinear least squares 
problems when some of the parameters appear linearly. However, setting up the 
calculations, that is, identifying which parameters are linear, is not a trivial
task.

Moreover, specifying a model to a solver should, ideally, use the same syntax across
solver tools. Unfortunately, R allows multiple approaches within different
modelling tools, and within `nls()` itself. This is highlighted by the case of
partially linear models.

@@

Within `nls()`, consider the following FOUR different specifications for the same problem, plus 
an intuitive choice, labelled `fm2a`, that does NOT work. In this failed
attempt, putting the `Asym` parameter in the model causes the `plinear` algorithm
to try to add another term to the model. We believe this is unfortunate, and would
like to see a consistent syntax. At the time of writing, we do
not foresee a resolution for this issue. In the example, we have NOT evaluated
the commands to save space.

@@

Indexed parameters -- those whose names have subscripts on some index variable -- 
are a useful idea to simplify model specification in some cases. However, they
are also a focus of trouble.

Here we use a truncated version of the example to save display space.

```{r nlsindx1}
## The muscle dataset in MASS is from an experiment on muscle
## contraction on 21 animals.  The observed variables are Strip
## (identifier of muscle), Conc (Cacl concentration) and Length
## (resulting length of muscle section).
if(! requireNamespace("MASS", quietly = TRUE)) stop("Need MASS pkg")
mm<- MASS::muscle[1:12,] # take only 1st few values of Strip (TRUNCATION OF EXAMPLE)
mm<-droplevels(mm) # remove unused levels after truncation
nlev <- nlevels(mm)
withAutoprint({
  ## The non linear model considered is
  ##       Length = alpha + beta*exp(-Conc/theta) + error
  ## where theta is constant. For now alpha and beta do NOT vary with Strip.
  with(mm, table(Strip)) # 2, 3 or 4 obs per strip
  nl <- nlevels(mm$Strip)
  ## We first use the plinear algorithm to fit an overall model,
  ## ignoring that alpha and beta might vary with Strip.
  musc.1 <- nls(Length ~ cbind(1, exp(-Conc/th)), mm,
                start = list(th = 1), algorithm = "plinear")
  summary(musc.1)

  ## Then we use nls' indexing feature for parameters in non-linear
  ## models to use the conventional algorithm to fit a model in which
  ## alpha and beta vary with Strip.  The starting values are provided
  ## by the previously fitted model.
  ## Note that with indexed parameters, the starting values must be
  ## given in a list (with names):
  ## ?? but why use b here AND in the new formula??
  b <- coef(musc.1)
  musc.2 <- nls(Length ~ a[Strip] + b[Strip]*exp(-Conc/th), data=mm,
                start = list(a = rep(b[2], nl), b = rep(b[3], nl), th = b[1]))
  summary(musc.2)
})
```

Running the example reveals that the answers for the parameters are NOT indexed. 
<!-- e.g. `coef(musc.2)` is a single level vector of parameters.  -->
That is, we do not see `a[1], a[2], a[3]` but `a1, a2, a3`. 
<!-- This is because the model -->
<!-- must integrate all the parameters because `th` is a common parameter across the index `Strip`. -->
This is no doubt because programming for indexed parameters is extremely challenging.
<!-- However, we believe the current structure could cause confusion and error, and  -->
<!-- propose an alternative approach below. -->

@@

@@

- to provide a packaged version of `nls()` (call it `nlsalt`) coded entirely in R 
  that matches the features in base R or what is packaged in `nlspkg` as described 
  in @PkgFromRbase22. The R version can either serve as the operational code or
  as a reference version to facilitate the maintenance of C, Fortran, or other 
  versions of the software.

@@

- to explain what we do, either in comments, examples, or separate maintainer documentation.
  Since we are complaining about the lack of explanatory material for the current
  code, we feel it is incumbent on us to provide such material for our own work, and if
  possible for the existing code.

@@

Nonlinear models, apart from those using the "plinear" algorithm, need all parameters specified.
That is, the model formula should be written down as it would be computed.
Thus a linear model (which is valid input for nonlinear estimation) should
be specified `y ~ a * x + b` rather than `y ~ x`. 
Moreover, we want to avoid the partially linear parameters appearing 
in the result object as `.lin1`, `.lin2`, etc., where the actual position 
of parameters in the model is not explicit.

### 2022-07-27 RefactoringNLS/RefactoringNLS.tex
Edits to formula section

The convergence criterion above, the principal termination control of
\texttt{nls()}, leaves out some possibilities that could be useful for
some problems. The package \texttt{nlsr} (John C Nash and Duncan Murdoch
(\protect\hyperlink{ref-nlsr-manual}{2019})) already offers both the
safeguarded relative offset test (\textbf{roffset}) as well as a
\textbf{small sum of squares} test (\textbf{smallsstest}) that compares
the latest evaluated sum of squared (weighted) residuals to \texttt{e4}
times the initial sum of squares, where
\texttt{e4\ \textless{}-\ (100*.Machine\$double.eps)\^{}4} is
approximately 2.43e-55.

@@

Specifying a model to a solver should, ideally, use the same syntax
across solver tools. Unfortunately, R allows multiple approaches within
different modelling tools, and within \texttt{nls()} itself.

@@

Within \texttt{nls()}, consider the following FOUR different
specifications for the same problem, plus an intuitive choice, labelled
fm2a, that does not work. In this failed attempt, putting the
\texttt{Asym} parameter in the model causes the \texttt{plinear}
algorithm to try to add another term to the model. We believe this is
unfortunate, and would like to see a consistent syntax. At the time of
writing, we do not foresee a resolution for this issue. In the example,
we have NOT evaluated the commands to save space.

@@

@@

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-nlsr-manual}{}}%
John C Nash, and Duncan Murdoch. 2019. \emph{{nlsr: Functions for
Nonlinear Least Squares Solutions}}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-NashWires22}{}}%
---------. 2022. {``Function Minimization and Nonlinear Least Squares in
r.''} \emph{WIREs Computational Statistics}, no. e1580.
https://doi.org/\url{https://doi.org/10.1002/wics.1580}.

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-HWtestthat11}{}}%
Wickham, Hadley. 2011. {``Testthat: Get Started with Testing.''}
\emph{The R Journal} 3: 5--10.
\url{https://journal.r-project.org/archive/2011-1/RJournal_2011-1_Wickham.pdf}.

\leavevmode\vadjust pre{\hypertarget{ref-HWdevtools21}{}}%
Wickham, Hadley, Jim Hester, Winston Chang, and Jennifer Bryan. 2021.
\emph{Devtools: Tools to Make Developing r Packages Easier}.
\url{https://CRAN.R-project.org/package=devtools}.

>>>>

The convergence criterion above, the principal termination control of
\texttt{nls()}, leaves out some possibilities that could be useful for
some problems. The package \texttt{nlsr}
((\protect\hyperlink{ref-nlsr-manual}{\textbf{nlsr-manual?}})) already
offers both the safeguarded relative offset test (\textbf{roffset}) as
well as a \textbf{small sum of squares} test (\textbf{smallsstest}) that
compares the latest evaluated sum of squared (weighted) residuals to
\texttt{e4} times the initial sum of squares, where
\texttt{e4\ \textless{}-\ (100*.Machine\$double.eps)\^{}4} is
approximately 2.43e-55.

@@

The variable projection method (Golub and Pereyra
(\protect\hyperlink{ref-Golub1973}{1973}), O'Leary and Rust
(\protect\hyperlink{ref-OlearyRust13}{2013})) is generally much more
effective then general approaches in finding good solutions to nonlinear
least squares problems when some of the parameters appear linearly.
However, setting up the calculations, that is, identifying which
parameters are linear, is not a trivial task.

Moreover, specifying a model to a solver should, ideally, use the same
syntax across solver tools. Unfortunately, R allows multiple approaches
within different modelling tools, and within \texttt{nls()} itself. This
is highlighted by the case of partially linear models.

@@

Within \texttt{nls()}, consider the following FOUR different
specifications for the same problem, plus an intuitive choice, labelled
\texttt{fm2a}, that does NOT work. In this failed attempt, putting the
\texttt{Asym} parameter in the model causes the \texttt{plinear}
algorithm to try to add another term to the model. We believe this is
unfortunate, and would like to see a consistent syntax. At the time of
writing, we do not foresee a resolution for this issue. In the example,
we have NOT evaluated the commands to save space.

@@

Indexed parameters -- those whose names have subscripts on some index
variable -- are a useful idea to simplify model specification in some
cases. However, they are also a focus of trouble.

@@

Here we use a truncated version of the example to save display space.

\begin{Shaded}
\begin{Highlighting}[]
\DocumentationTok{\#\# The muscle dataset in MASS is from an experiment on muscle}
\DocumentationTok{\#\# contraction on 21 animals.  The observed variables are Strip}
\DocumentationTok{\#\# (identifier of muscle), Conc (Cacl concentration) and Length}
\DocumentationTok{\#\# (resulting length of muscle section).}
\ControlFlowTok{if}\NormalTok{(}\SpecialCharTok{!} \FunctionTok{requireNamespace}\NormalTok{(}\StringTok{"MASS"}\NormalTok{, }\AttributeTok{quietly =} \ConstantTok{TRUE}\NormalTok{)) }\FunctionTok{stop}\NormalTok{(}\StringTok{"Need MASS pkg"}\NormalTok{)}
\NormalTok{mm}\OtherTok{\textless{}{-}}\NormalTok{ MASS}\SpecialCharTok{::}\NormalTok{muscle[}\DecValTok{1}\SpecialCharTok{:}\DecValTok{12}\NormalTok{,] }\CommentTok{\# take only 1st few values of Strip}
\FunctionTok{str}\NormalTok{(mm)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## 'data.frame':    12 obs. of  3 variables:
##  $ Strip : Factor w/ 21 levels "S01","S02","S03",..: 1 1 1 1 2 2 2 2 3 3 ...
##  $ Conc  : num  1 2 3 4 1 2 3 4 0.25 0.5 ...
##  $ Length: num  15.8 20.8 22.6 23.8 20.6 26.8 28.4 27 7.2 15.4 ...
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{mm}\OtherTok{\textless{}{-}}\FunctionTok{droplevels}\NormalTok{(mm)}
\FunctionTok{str}\NormalTok{(mm)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## 'data.frame':    12 obs. of  3 variables:
##  $ Strip : Factor w/ 3 levels "S01","S02","S03": 1 1 1 1 2 2 2 2 3 3 ...
##  $ Conc  : num  1 2 3 4 1 2 3 4 0.25 0.5 ...
##  $ Length: num  15.8 20.8 22.6 23.8 20.6 26.8 28.4 27 7.2 15.4 ...
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{nlev }\OtherTok{\textless{}{-}} \FunctionTok{nlevels}\NormalTok{(mm)}
\FunctionTok{withAutoprint}\NormalTok{(\{}
  \DocumentationTok{\#\# The non linear model considered is}
  \DocumentationTok{\#\#       Length = alpha + beta*exp({-}Conc/theta) + error}
  \DocumentationTok{\#\# where theta is constant but alpha and beta may vary with Strip.}
  \FunctionTok{with}\NormalTok{(mm, }\FunctionTok{table}\NormalTok{(Strip)) }\CommentTok{\# 2, 3 or 4 obs per strip}
\NormalTok{  nl }\OtherTok{\textless{}{-}} \FunctionTok{nlevels}\NormalTok{(mm}\SpecialCharTok{$}\NormalTok{Strip)}
  \DocumentationTok{\#\# We first use the plinear algorithm to fit an overall model,}
  \DocumentationTok{\#\# ignoring that alpha and beta might vary with Strip.}
\NormalTok{  musc}\FloatTok{.1} \OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(Length }\SpecialCharTok{\textasciitilde{}} \FunctionTok{cbind}\NormalTok{(}\DecValTok{1}\NormalTok{, }\FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{Conc}\SpecialCharTok{/}\NormalTok{th)), mm,}
                \AttributeTok{start =} \FunctionTok{list}\NormalTok{(}\AttributeTok{th =} \DecValTok{1}\NormalTok{), }\AttributeTok{algorithm =} \StringTok{"plinear"}\NormalTok{)}
  \FunctionTok{summary}\NormalTok{(musc}\FloatTok{.1}\NormalTok{)}

  \DocumentationTok{\#\# Then we use nls\textquotesingle{} indexing feature for parameters in non{-}linear}
  \DocumentationTok{\#\# models to use the conventional algorithm to fit a model in which}
  \DocumentationTok{\#\# alpha and beta vary with Strip.  The starting values are provided}
  \DocumentationTok{\#\# by the previously fitted model.}
  \DocumentationTok{\#\# Note that with indexed parameters, the starting values must be}
  \DocumentationTok{\#\# given in a list (with names):}
  \DocumentationTok{\#\# ?? but why use b here AND in the new formula??}
\NormalTok{  b }\OtherTok{\textless{}{-}} \FunctionTok{coef}\NormalTok{(musc}\FloatTok{.1}\NormalTok{)}

\NormalTok{  musc}\FloatTok{.2} \OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(Length }\SpecialCharTok{\textasciitilde{}}\NormalTok{ a[Strip] }\SpecialCharTok{+}\NormalTok{ b[Strip]}\SpecialCharTok{*}\FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{Conc}\SpecialCharTok{/}\NormalTok{th), }\AttributeTok{data=}\NormalTok{mm,}
                \AttributeTok{start =} \FunctionTok{list}\NormalTok{(}\AttributeTok{a =} \FunctionTok{rep}\NormalTok{(b[}\DecValTok{2}\NormalTok{], nl), }\AttributeTok{b =} \FunctionTok{rep}\NormalTok{(b[}\DecValTok{3}\NormalTok{], nl), }\AttributeTok{th =}\NormalTok{ b[}\DecValTok{1}\NormalTok{]))}
  \FunctionTok{summary}\NormalTok{(musc}\FloatTok{.2}\NormalTok{)}
\NormalTok{\})}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## > with(mm, table(Strip))
## Strip
## S01 S02 S03 
##   4   4   4 
## > nl <- nlevels(mm$Strip)
## > musc.1 <- nls(Length ~ cbind(1, exp(-Conc/th)), mm, start = list(th = 1), 
## +     algorithm = "plinear")
## > summary(musc.1)
## 
## Formula: Length ~ cbind(1, exp(-Conc/th))
## 
## Parameters:
##       Estimate Std. Error t value Pr(>|t|)    
## th       0.624      0.222    2.81   0.0203 *  
## .lin1   25.684      1.441   17.82  2.5e-08 ***
## .lin2  -26.631      6.147   -4.33   0.0019 ** 
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## Residual standard error: 2.96 on 9 degrees of freedom
## 
## Number of iterations to convergence: 5 
## Achieved convergence tolerance: 4.71e-07
## 
## > b <- coef(musc.1)
## > musc.2 <- nls(Length ~ a[Strip] + b[Strip] * exp(-Conc/th), data = mm, 
## +     start = list(a = rep(b[2], nl), b = rep(b[3], nl), th = b[1]))
## > summary(musc.2)
## 
## Formula: Length ~ a[Strip] + b[Strip] * exp(-Conc/th)
## 
## Parameters:
##    Estimate Std. Error t value Pr(>|t|)    
## a1  22.9277     0.5154   44.49  1.1e-07 ***
## a2  27.8606     0.5150   54.09  4.1e-08 ***
## a3  28.3426     1.0771   26.31  1.5e-06 ***
## b1 -44.1558    12.3900   -3.56    0.016 *  
## b2 -43.8084    12.3148   -3.56    0.016 *  
## b3 -32.8421     2.1553  -15.24  2.2e-05 ***
## th   0.5548     0.0808    6.86    0.001 ** 
## ---
## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
## 
## Residual standard error: 0.798 on 5 degrees of freedom
## 
## Number of iterations to convergence: 6 
## Achieved convergence tolerance: 1.94e-06
\end{verbatim}

@@

\leavevmode\vadjust pre{\hypertarget{ref-Golub1973}{}}%
Golub, G. H., and V. Pereyra. 1973. {``The Differentiation of
Pseudo-Inverses and Nonlinear Least Squares Problems Whose Variables
Separate.''} \emph{SIAM Journal of Numerical Analysis} 10 (2): 413--32.

@@

@@

\leavevmode\vadjust pre{\hypertarget{ref-NashWires22}{}}%
---------. 2022. {``{Function minimization and nonlinear least squares
in R}.''} \emph{WIREs Computational Statistics}, no. e1580.
https://doi.org/\url{https://doi.org/10.1002/wics.1580}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-OlearyRust13}{}}%
O'Leary, Dianne P., and Bert W. Rust. 2013. {``Variable Projection for
Nonlinear Least Squares Problems.''} \emph{Computational Optimization
and Applications} 54 (3): 579--93.

@@

\leavevmode\vadjust pre{\hypertarget{ref-HWtestthat11}{}}%
Wickham, Hadley. 2011. {``{testthat: Get Started with Testing}.''}
\emph{The R Journal} 3: 5--10.
\url{https://journal.r-project.org/archive/2011-1/RJournal_2011-1_Wickham.pdf}.

\leavevmode\vadjust pre{\hypertarget{ref-HWdevtools21}{}}%
Wickham, Hadley, Jim Hester, Winston Chang, and Jennifer Bryan. 2021.
\emph{{devtools: Tools to Make Developing R Packages Easier}}.
\url{https://CRAN.R-project.org/package=devtools}.
